CAD and Routing Architecture for Interposer-Based Multi-FPGA Systems

### 2014-12-07 doc/journal/fpgapaperfinal/fpga2014paper.tex
updating files to the latest version before changing template from ACM to IEEETran
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4520 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

\section{Architecture models}
\label{archSection}
%   - Define parameters
To properly model a silicon interposer FPGA, the popular FPGA exploration toolset, Verilog-to-Routing (VTR)~\cite{vtr2012}, was used. The logic synthesis portions of the flow (ODIN II and  ABC) were left untouched, while the placement and routing portion of the flow (VPR~\cite{betz1997vpr}) was modified to model and optimize for interposer-based FPGAs. The modifications were made in such a way that they require no changes to any of the input files, so one can experiment with interposer-based FPGAs with any existing benchmark circuits and any existing VPR-format FPGA architecture description simply by specifying appropriate command-line parameters.

Three parameters were added to VPR: \textit{\% wires cut}, \textit{increased delay} and \textit{number of cuts}. These three parameters describe the interposer portion of the 2.5D FPGA as detailed below.

@@

%	- number of cuts
\subsection{Number of cuts}
\textit{Number of cuts} describes how many cuts were made to the interposer-based FPGA versus a monolithic die. If \textit{number of cuts} equals 1 then there are 2 dies, and so on. We investigate values of this parameter between 1 and 3 (between 2 and 4 dies), reflecting commercial practice: the Virtex 7 family has members with \textit{number of cuts} $=$ 2 and 3. Figure \ref{fig:fpga} shows a sample architecture with one cutline.

%   - (briefly) explain how it was put in VPR
\subsection{Implementation}
To model an interposer-based FPGA in VPR the Routing Resource Graph (\textit{rrgraph}) must be modified. The rrgraph is the data structure that defines all the available routing wires and switches in the FPGA, as well the delay of each. Given a suitable rrgraph, the VPR router can implement circuits in the desired FPGA, and the VPR timing analyzer can estimate their delay.

The presence of multiple dice in an interposer-based FPGA was modeled by creating horizontal cuts in the FPGA, which are equally spaced vertically. Every wire that crosses one of these cuts in fact passes through the interposer. In the experiments below, we use an architecture that has only unidirectional wires~\cite{unidirectional}, as these are the dominant routing architecture in modern FPGAs. Such wires can only be driven at their beginning. To control the wiring capacity between dies \textit{\% wires cut} wires of each channel segment crossing a cutline have their connections to wires and block inputs on the die opposite to their starting point removed, and the wires which aren't cut have \textit{increased delay} added to their connections which cross the cutline. Note that the combination of a unidirectional routing architecture and a silicon interposer results in some blocks near the cutline having reduced routing connectivity as some of the routing wires driving their inputs are disconnected from the (single) wire driver, when that driver is on the other side of the cutline and the wire is one of the cut wires.

Figure \ref{fig:implementation} illustrates the approach used.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{implementation.eps}
\caption{Illustration of the approach used to simulate crossing the interposer. The red dashed line indicates the interposer, the black arrows indicate the direction of the wires. Blue arrows indicate normal connections from the wires to the CLBs. The red arrows indicate removed connections and green arrows indicate connections with increased delay.}
\label{fig:implementation}
\end{figure}

For all experiments in this paper, the remaining architecture parameters of the FPGA are taken from the "flagship" architecture of the VTR project (\textit{k6\_frac\_N10\_mem32K\_40nm.xml}). The parameters of this architecture are in line with both current commercial FPGAs and academic research into best practices. It consists of logic clusters with 10 fracturable 6-LUTs per block (N=10, k=6), and also includes 32kb RAM blocks and DSP blocks configurable to perform 9x9, 18x18 or 36x36 multipliers. The delay values in the architecture are taken from 40 nm circuit simulations and 40 nm commercial FPGAs. It uses unidirectional routing, with all wire segments having length L = 4.

\section{CAD Enhancements}
\label{cadSection}

Once the routing resource graph was modified as detailed in Section \ref{archSection}, the VPR router adapted automatically to the interposer architecture. Placement, however, is crucial in mitigating the impact of the reduced wiring and increased delay when crossing between dice; a good placement should minimize the number of signals crossing between dice, particularly time-critical ones. We investigated several enhancements to VPR's placement cost function to improve result quality.

VPR uses 2 different costs as the metrics for its placer algorithm: the timing cost and the bounding box (wiring) cost. The usual VPR timing cost is a (criticality-weighted) summation of the estimated delay (given the current placement) of every connection required by the circuit~\cite{timing2000}:

\begin{equation} \label{eq:timing_eq_full}
\begin{split}
Timing\_Cost = \sum_{\forall i, j \subset circuit} delay(\Delta x_{ij},\Delta y_{ij}) \times \\ 
criticality(i,j)
\end{split}
\end{equation}

where \textit{ij} denotes a connection from block i to block j that exists in the circuit netlist.
The bounding box cost estimates the amount of wiring required for a net, based on the number of pins and size of the net's bounding box. VPR's original formulation is~\cite{betz1997vpr}:

where $bb_x(n)$ and $bb_y(n)$ are the dimensions of the bounding box of net \textit{n} in the x and y directions, respectively. $avg\_chanx\_W$ and $avg\_chany\_W$ are the average x-directed and y-directed channel widths over this bounding box. Finally, $q(n)$ is a function obtained from \cite{icann} which models the fact that bounding boxes underpredict the required routing for high fanout nets. $q(n)$ slowly increases with the fanout of net \textit{n}, rising from 1 for nets with 3 or fewer terminals to 2.79 for nets with 50 terminals.

We modified both the placer's bounding box and timing costs to account for the increased latency to cross the interposer and for the reduced wire capacity close to the cutlines, as the wire availability becomes more sparse.

%   - Delay
\subsection{Placer timing cost}
The standard placer timing cost in VPR assumes that the FPGA is homogeneous and consequently the delay between 2 points $(x_1,y_1)$ and $(x_2,y_2)$ only depends on $(\Delta x,\Delta y)$. This is obviously not true for interposer-based FPGAs, as the cutlines make them heterogeneous in the \textit{y} direction.

To solve this problem and improve the quality of the results, an extra term was added to the delay function. The delay function becomes:

@@

%   - Wiring / cut cost
\subsection{Placer wiring cost}
VPR's wiring cost also considers the FPGA to be homogeneous, and uses only the number of nets, the size of the net's bounding box and the average number of wires per channel to calculate the cost. Thus, to account for the reduced connectivity near the cutlines an extra cost term, \textit{cut\_cost}, was created. This new cost is added to (\ref{eq:bbcost}) to create the total wiring cost.

@@

We tested several different \textit{cut\_cost} formulations, as well as different weighting (C values) for each. For all of the formulations the variable $C'$ was defined as:

\begin{equation} \label{eq:cprime}
\begin{split}
C' = \frac{C \times ratio\_wires\_cut}{avg\_chany\_W(n)}
\end{split}
\end{equation}

where $ratio\_wires\_cut$ is the ratio of wires cut at the cutline. This formulation of $C'$ ensures that when we choose a \textit{C} value of 1, the new term cost term is of roughly the same magnitude as $wire\_cost_{orig}$ in (\ref{eq:bbcost}) and that it is weighted more heavily for interposer architectures in which the wiring between dice is more scarce.

The first \textit{cut\_cost} term we tested was:

\begin{equation} \label{eq:cost0}
cut\_cost = \sum_{n=1}^{N_{nets}} C' \times times\_crossed(n)
\end{equation}

which penalizes the net according to the number of cutlines the bounding box spans; this directly penalizes each signal crossing between dice. Surprisingly this extra term did not result in significant improvements to the quality of results and in some cases even made them worse. We believe this is due to the discontinuous, non-smooth, nature of this cost function -- it does not guide placement optimization well as a bounding box shrinks toward a cutline. Placement changes that make bounding boxes "almost not cross" a cutline are not given any gain; only a sudden change in the placement that moves a bounding box entirely to one side or the other of the cutline yields a cost reduction. This insight led us to the other proposed functions, which give gradual gains as bounding boxes change in their size and come closer to avoiding a cutline crossing.

The other tested terms were:

%add C' x times_crossed(i)

\begin{equation} \label{eq:cost1}
cut\_cost = \sum_{n=1}^{N_{nets}} C' \times bbWidth(n) \times times\_crossed(n)
\end{equation}

@@

\begin{equation} \label{eq:cost4}
\begin{split}
cut\_cost = \sum_{n=1}^{N_{nets}} C' \times minDist(i) \times times\_crossed(n) +\\
C' \times times\_crossed(n)
\end{split}
\end{equation}

\begin{equation} \label{eq:cost5}
cut\_cost = \sum_{n=1}^{N_{nets}} C' \times bbHeight(n)
\end{equation}

where \textit{bbWidth(n)} and \textit{bbHeight(n)} are the width and height of the bounding box of the net \textit{n}, respectively, and \textit{minDist(n)} is the minimum distance from the top or bottom of the bounding box to a cutline.

%   - Effectiveness / results
\subsection{Effectiveness of the enhancements}
\label{sec:CADeffect}

We used VPR 7.0, with the enhancements we detailed above, and the architecture file \textit{k6\_frac\_N10\_mem32K\_40nm} in the experiments below. All experiments targeted the smallest FPGA (with number of rows equal to number of columns) which could accomodate a benchmark circuit; this represents a very full FPGA with little white space left, and hence presents a difficult case to an interposer-based FPGA as no die can be left mostly empty.

Figures \ref{fig:minW_constant_sweep} and \ref{fig:areadelay_constant_sweep} show the efficacy of the five placer wiring cost modifications as their weight, \textit{C} varies. Each point in the graphs is the geometric mean of the results with \textit{\% wires cut} = 60 and 80, and the results for 60\% and 80\% are themselves geometric means of the results of 6 circuits from the VTR benchmark suite~\cite{vtr2012}:  \textit{stereovision0}, \textit{stereovision1}, \textit{mkDelayWorker32B}, \textit{mkSMAdapter4B}, \textit{or1200} and \textit{blob\_merge}. These values for \textit{\% wires cut} were chosen because they in the likely range of wires which can cross the interposer without exceeding the microbump capacity, as described in Section~\ref{virtex7section}. The \textit{area-delay product} is the product of the \textit{minimum channel width} and the \textit{critical path delay}. To obtain the \textit{critical path delay} the circuits were run with a low stress routing with a channel width, $W = 1.3 \times minW$, where $minW$ is the minimum channel width for which the circuit is still routable.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{minW_constant_sweep_log.eps}
\caption{Minimum channel width vs. weighting for different \textit{cut\_cost} terms.}
\label{fig:minW_constant_sweep}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{areadelay_constant_sweep_log.eps}
\caption{Area-delay product vs. weighting for different \textit{cut\_cost} terms.}
\label{fig:areadelay_constant_sweep}
\end{figure}

\begin{table}[htbp]
\begin{tabular}{|r|r|r|r|r|}
\hline
\multicolumn{1}{|l|}{Term} & \multicolumn{1}{l|}{Best C} & \multicolumn{1}{l|}{minW} & \multicolumn{1}{l|}{crit\_path(ns)} & \multicolumn{1}{l|}{Area-delay} \\ \hline \hline
None & - & 124.636 & 9.844 & 1226.955 \\ \hline
\ref{eq:cost0} & 0.3 & 122.929 & 9.413 & 1157.09 \\ \hline
\ref{eq:cost1} & 5 & 114.444 & 9.565 & 1094.68 \\ \hline
\ref{eq:cost2} & 1 & 107.303 & 9.285 & 996.26 \\ \hline
\ref{eq:cost4} & 1 & 113.381 & 9.280 & 1052.16 \\ \hline
\ref{eq:cost5} & 5 & 112.703 & 9.841 & 1109.11 \\ \hline
\end{tabular}
\caption{Best weighting and performance for each \textit{cut\_cost} term.}
\label{table:constant_sweep}
\end{table}

The best \textit{C} value and the resulting performance for each term is summarized in Table~\ref{table:constant_sweep}. Note that cost term (\ref{eq:cost2}) with weighting $C = 1.0$ has the best performance; this is the configuration used in the rest of our experiments.

We believe the key to the good performance of this cost function is that it produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline and hence are penalized equally by (\ref{eq:cost0}), while bounding box (c) does not cross a cutline and hence is not penalized. However, bounding box (b) is mostly on the lower side of the cutline; it is more likely that later smaller placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function terms (\ref{eq:cost2}) and (\ref{eq:cost4}) will both penalize bounding box (a) more than bounding box (b) to guide placement to gradually move bounding boxes to one side or the other of a cutline. Figure \ref{fig:areadelay_constant_sweep} shows that these are the two best performing \textit{cut\_cost} terms.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{bb.eps}
\caption{Illustration of three different scenarios for the wiring cost. The dashed green box shows a case where \textit{(a)} it crosses the interposer, the dotted black box shows a case where \textit{(b)} it barely crosses the cutline, and the solid blue one shows a case where \textit{(c)} the bounding box does not span the cutline.}
\label{fig:bb_illustration}
\end{figure}

Figure \ref{fig:comparison_minW} shows the minimum channel width for different values of \textit{\% wires cut} when there are no optimizations and when \textit{cut\_const} (\ref{eq:cost2}) is used. % comment about impact vs % wires cut
Figure \ref{fig:comparison_bars} compares result quality with both enhanced timing and wiring cost functions. These experiments were run with term (\ref{eq:cost2}), and are the geometric means of the runs with \textit{\% wires cut = 60, 70, 80}.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{comparison_bars_relative.eps}
\caption{Performance of different placer optimizations, relative to the original VPR results. The area-delay product is calculated as the product $minW \times critical path delay.$}
\label{fig:comparison_bars}
\end{figure}

The graphs show that combining the timing with the bounding box optimizations led to an average of 20\% improvement on the area-delay product and the most significant contribution was made by incorporation of \textit{cut\_cost} into the wire cost function.

%----------------------------------------------------
\section{Architecture results}
\label{resultsSection}

Using the best CAD settings found in Section~\ref{cadSection} we analyzed the impact of three key architecture parameters:  \textit{\% wires cut}, \textit{delay increase} and \textit{number of cuts}. All the experiments were run with the eight largest circuits from the VTR benchmark~\cite{vtr2012}, namely: \textit{bgm}, \textit{LU8PEEng}, \textit{LU32PEEng}, \textit{mcml}, \textit{mkDelayWorker32B}, \textit{stereovision0}, \textit{stereovision1} and \textit{stereovision2}. The size of these circuits ranges from $9,100$ to $153,000$ primitives (LUTs, FFs, etc.), with an average size of $52,600$ primitives. All results are the geometric mean over all circuits for a given interposer architecture.

>>>>

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture models}
\label{archSection}
%   - Define parameters
To properly model a silicon interposer FPGA, we use version 7.0 of the popular FPGA exploration toolset, Verilog-to-Routing (VTR)~\cite{luu2014vtr}. The logic synthesis portions of the flow (ODIN II and  ABC) are left untouched, while we modify the placement and routing portion of the flow (VPR~\cite{betz1997vpr}) to model and optimize for interposer-based FPGAs. The modifications are made in such a way that they require no changes to any of the input files, so one can experiment with interposer-based FPGAs with any existing benchmark circuits and any existing VPR-format FPGA architecture description simply by specifying appropriate command-line parameters.

We add the following parameters to VPR: \textit{number of cuts}, \textit{\% wires cut}, \textit{increased delay}, and \textit{interposer-routing interface}. These parameters describe the interposer portion of the 2.5D FPGA as detailed below.

%	- number of cuts
\subsection{Number of cuts}
\textit{Number of cuts} describes how many cuts are made to the interposer-based FPGA versus a monolithic die. If \textit{number of cuts} equals 1 then there are 2 dice, and so on. We investigate values of 1 and 3 (2 and 4 dice). The Virtex 7 family has members with \textit{number of cuts} $=$ 2 and 3. Figure \ref{fig:fpga} shows a sample architecture with one cutline.

@@

%   - Interposer-Routing interface
\subsection{Interposer-Routing interface}
\label{sec:interposer-routing-interface}
In a simple architecture, the vertical routing wires that can not connect to an interposer wire are dangling. Since interposer wires are scarce, we want to be able to study architectures in which the interposer wires are more accessible. To do so, we have enhanced the router to support the following interposer-routing interface options.

\begin{enumerate}[label=(\alph*)]
\item Fan-in Transfer for Interposer Wires\\
For an architecture with \textit{\% wires cut} $=0$, all vertical routing wires have a corresponding interposer wire that connects them to the adjacent die. Practical interposers, however, can only accommodate a fraction of all vertical routing wires. When \textit{fanin transfer} option is \textit{on}, we place a multiplexer at the input of each available interposer wire and connect some adjacent ``dangling" vertical routing wires to the multiplexer. This allows for additional flexibility in routing by ensuring that \textit{every} vertical routing wire could \textit{potentially} drive an interposer wire through a multiplexer. As an example, if \textit{\% wires cut} $=75\%$, 3 of every 4 interposer wires are cut. Therefore, we place a 4-to-1 multiplexer at the input of the one interposer wire that is not cut, and we connect the 3 vertical routing wires that do not have a corresponding interposer wire to the multiplexer.

\item Fan-out Transfer for Interposer Wires\\
When \textit{fanout transfer} option is \textit{on}, we allow the interposer wire output to drive not only its own corresponding vertical wire at the other side of the cut, but also the other ``dangling" vertical routing wires that are not driven by anything because their corresponding interposer wires are cut.

\item Bidirectional Interposer Wires\\
Dominant routing architectures in modern FPGAs use unidirectional wires~\cite{unidirectional, lewis2005stratix}. Such wires can only be driven at one point, normally their start point. When this option is \textit{on}, we allow interposer wires to be driven from one end or the other by adding tri-state buffers at both ends.
\end{enumerate}

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{architecture_params.eps}
\caption{For a channel width of 4 and 75\% wires cut, only 1 interposer wire exists (in green) and 3 interposer wires are cut (in the red). (a)~Fan-in Transfer for Interposer Wires allows red routing wires in die 1  to drive a mux that allows them to cross the interposer layer. (b)~Fan-out Transfer for Interposer Wires allows the output of the interposer wire to drive wires in other tracks in die 2.}
\label{fig:architecture_params}
\end{figure}

These options are independent and can be combined to achieve more flexible routing. Enabling any of these options requires adding specific circuitry (such as multiplexers) on the dice which causes very small area and delay increase that we neglect in our experiments.

%   - (briefly) explain how it was put in VPR
\subsection{Implementation}
To model an interposer-based multi-FPGA system in the CAD tool, we use the existing VPR infrastructure for monolithic FPGAs and make several modifications to it.

First, The presence of multiple dice in an interposer-based FPGA is modeled by creating horizontal cuts in the FPGA. The \textit{number of cuts} is specified by the user, and we use that to determine the y-location of the cuts. The cutlines ``snap to the grid" such that the height of all dice are equal and no cutline cuts through a block. For instance, a RAM block of height 4 must remain a unit and can not be sliced by a cutline.

Second, knowing the location of the cutlines, the placer uses the \textit{(x,y)} coordinate of any pin or block to determine the die to which it belongs. The placer uses this information as well as \textit{interposer delay} to better decide where to place each netlist block in order to minimize the cost of inter-die signal transfer as described in Section \ref{cad_enh_placer_subsection}.

Third, the Routing Resource Graph (\textit{rrgraph}) must be modified. The rrgraph is the data structure that defines all the available routing wires (\textit{rrnodes}) and switches in the FPGA, as well as the delay of each~\cite{betz1999architecture}. Given a suitable rrgraph, the VPR router can implement circuits in the desired FPGA, and the VPR timing analyzer can estimate their delay. In the experiments below, we use an architecture that has only unidirectional wires.

Previous work by Pereira et al \cite{hahn2014cad} modeled the interposer indirectly: they did not introduce new rrnodes in the rrgraph to represent the interposer wires. Rather, they found the rrnodes that cross a cutline. Then they either removed all outgoing edges of these rrnodes (if the routing node was part of the \textit{\% wires cut}), or, they increased the switch delay of the rrnode fan-outs (if the wire was not cut) to account for the additional interposer delay.

While this is a reasonable method to model such architectures, it makes it difficult to experiment with different routing-interposer interfaces such as placing input muxes for interposer wires. Therefore, we modify the rrgraph differently to more accurately reflect the interposer layer directly:

\begin{figure}[!t]
\subfigure[]{\includegraphics[width=0.24\linewidth]{a.eps}}
\subfigure[]{\includegraphics[width=0.24\linewidth]{b.eps}}
\subfigure[]{\includegraphics[width=0.24\linewidth]{a_prime.eps}}
\subfigure[]{\includegraphics[width=0.24\linewidth]{b_prime.eps}}
\caption{(a) Regular routing wire connectivity. (b) Replacing the long wire with two wires and connecting them via an interposer wire. (c) rrgraph for regular connectivity. (d) rrgraph after modifications.}
\label{fig:rrgraph_ops}
\end{figure}

\begin{figure*}[!thpb]
\subfigure[]{\includegraphics[width=0.6\textwidth]{gui_1.eps}}
\subfigure[]{\includegraphics[width=0.4\textwidth]{gui_2.eps}}
\caption{VPR Graphical User Interface. (a) Full chip view. (b) Connectivity at the cutline. Due to the limited signal capacity of the interposer, some vertical routing wires are left dangling.}
\label{fig:vpr_interposer_gui}
\end{figure*}

For every vertical routing wire in the channel, we add a new routing node (``\textit{interposer node}") into the rrgraph. Then, we determine rrnodes (e.g. vertical routing wires) that cross a cutline in the rrgraph, and we replace them with 2 rrnodes: one that is above the cutline and one that is below the cutline. We properly transfer fan-ins and fan-outs of the original node onto the two newly created nodes and we connect the two nodes using an interposer node. Finally, we remove \textit{all} edges from \textit{\%wires cut} percent of interposer nodes and increase the switch delay of output edges of the remaining interposer nodes by \textit{increased delay}. Figure~\ref{fig:rrgraph_ops} shows an example of these operations.

As illustrated in Figure~\ref{fig:vpr_interposer_gui}, no vertical routing wires cross a cutline boundary and a fraction of them connect to other dies via interposer wires.

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{CAD Enhancements}
\label{cadSection}

\subsection{Placer optimization}
\label{cad_enh_placer_subsection}
Placement is crucial in mitigating the impact of the reduced wiring and increased delay when crossing between dice; a good placement should minimize the number of signals crossing between dice, particularly time-critical ones. VPR uses 2 different costs as the metrics for its placer algorithm: the timing cost and the bounding box (wiring) cost. We modified the placer's bounding box and timing costs to account for the increased latency to cross the interposer and for the reduced wire capacity close to the cutlines, as the wire availability becomes more sparse.

%   - Delay
\subsubsection{Placer timing cost}
The usual VPR timing cost is a (criticality-weighted) summation of the estimated delay (given the current placement) of every connection required by the circuit~\cite{timing2000}:

\begin{equation} \label{eq:timing_eq_full}
\begin{split}
Timing\_Cost = \sum_{\forall i, j \subset circuit} delay(\Delta x_{ij},\Delta y_{ij}) \times \\ 
criticality(i,j)
\end{split}
\end{equation}
where \textit{ij} denotes a connection from block i to block j that exists in the circuit netlist.

This cost function assumes that the FPGA is homogeneous and consequently the delay between 2 points $(x_1,y_1)$ and $(x_2,y_2)$ only depends on $(\Delta x,\Delta y)$. This is obviously not true for interposer-based FPGAs, as the cutlines make them heterogeneous in the \textit{y} direction. To solve this problem and improve the quality of the results, we add an extra term to the delay function:

@@

%   - Wiring / cut cost
\subsubsection{Placer wiring cost}
The bounding box cost estimates the amount of wiring required for a net, based on the number of pins and size of the net's bounding box. VPR's original formulation is~\cite{betz1997vpr}:

where $bb_x(n)$ and $bb_y(n)$ are the dimensions of the bounding box of net \textit{n} in the x and y directions, respectively. $avg\_chanx\_W$ and $avg\_chany\_W$ are the average x-directed and y-directed channel widths over this bounding box. Finally, $q(n)$ is a function obtained from \cite{icann} which models the fact that bounding boxes underpredict the required routing for high fanout nets. $q(n)$ slowly increases with the fanout of net \textit{n}.

VPR's wiring cost also considers the FPGA to be homogeneous, and uses only the number of nets, the size of the net's bounding box and the average number of wires per channel to calculate the cost. Thus, to account for the reduced connectivity near the cutlines we create an extra cost term, \textit{cut\_cost}. This new cost is added to (\ref{eq:bbcost}) to create the total wiring cost.

@@

In~\cite{hahn2014cad}, Pereira et al tested several different \textit{cut\_cost} formulations, as well as different weighting for each. They found that the following \textit{cut\_cost} results in the best quality of results in terms of minimum routable channel width and critical path delay.

@@

where 
\begin{equation} \label{eq:cprime}
\begin{split}
C' = \frac{ratio\_wires\_cut}{avg\_chany\_W(n)}
\end{split}
\end{equation}

\textit{bbHeight(n)} is the height of the bounding box of the net \textit{n} and $ratio\_wires\_cut$ is the ratio of wires cut at the cutline. This formulation of $C'$ ensures that the new cost term is of roughly the same magnitude as $wire\_cost_{orig}$ in (\ref{eq:bbcost}) and that it is weighted more heavily for interposer architectures in which the wiring between dice is more scarce.

This cost function performs better than several others as shown in \cite{hahn2014cad}, and we believe the key to the good performance of this cost functions is that \textit{cut\_cost} produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline, but (b) is penalized less by (\ref{eq:cost2}) because bounding box (b) is mostly on the lower side of the cutline; it is more likely that later placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function (\ref{eq:cost2}) penalize bounding box (a) more than (b) and (b) more than (c) to guide placement to gradually move bounding boxes to one side or the other of a cutline.

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{bb_new.eps}
\caption{Illustration of three different scenarios for the wiring cost. The dashed green box shows a case where \textit{(a)} it crosses the interposer, the dotted black box shows a case where \textit{(b)} it barely crosses the cutline, and the solid blue one shows a case where \textit{(c)} the bounding box does not span the cutline.}
\label{fig:bb_illustration}
\end{figure}

\subsection{Router optimization}
\label{sec:router_optimizations}
Once the Routing Resource Graph is modified as detailed in Section \ref{archSection}, the VPR router adapted automatically to the interposer architecture. The router starts at a net source node and expands by discovering adjacent routing nodes until it reaches the destination node. At any point, the router knows the cost of the path taken to the current point and it uses a \textit{lookahead cost function} to estimate the resource and delay cost from the current node to the destination. Using this lookahead cost, the router can properly sort routing alternatives. The lookahead cost function in VPR is not aware of the interposer, and therefore it takes a long time to converge to a routing decision. To make the lookahead aware of the interposer, we modify the lookahead cost function by adding terms that account for extra \textit{interposer delay} and extra \textit{interposer nodes} that may exist on the route from the current node to the destination node.

The VPR router uses the following to calculate a cost for a routing node \textit{r} on the route from \textit{i} to \textit{j}:
\begin{equation} \label{eq:routing_cost}
\begin{split}
Cost(r) = Criticality(i,j) \times Delay_{Elmore}(r, topology) + \\ [1-Criticality(i,j)] \times Congestion(r)
\end{split}
\end{equation}

Given this cost function, the router's lookahead algorithm determines the total cost of the route from \textit{i} to \textit{j} when routing node \textit{r} is used:

\begin{equation} \label{eq:lookahead_cost}
\begin{split}
Total\_Cost(r) = \sum_{\forall l \in RT(i,r)} Cost(l) + \alpha \times ExpectedCost(r,j)
\end{split}
\end{equation}

where \textit{RT(i,r)} is the set of all routing nodes used so far from \textit{i} to \textit{r} and $\alpha$ is a constant number that determines the aggressiveness of the directed search. $\alpha$ is 1.2 by default in VPR. VPR's original $ExpectedCost(n,j)$ function is~\cite{betz1999architecture}:

\begin{equation} \label{eq:expected_routing_cost}
\begin{split}
ExpectedCost(r,j) = Criticality_{fac} \times T_{delay} + \\(1 - Criticality_{fac}) \times ExpectedCongestionCost(r,j);
\end{split}
\end{equation}

$T_{delay}$ is an estimate of the delay that the router predicts it will see on the route from \textit{r} to \textit{j}. Since (\ref{eq:expected_routing_cost}) assumes a homogeneous FPGA, $T_{delay}$ does not include the additional delay some routes experience by going through the interposer. Thus, for interposer-based architectures, we replace $T_{delay}$ in (\ref{eq:expected_routing_cost}) with $T^\prime_{delay}$ to include the delay of expected interposer hops from \textit{r} to \textit{j}.
\begin{equation} \label{eq:tdelay}
\begin{split}
T^\prime_{delay} = T_{delay} + T_{interposer} \times ExpectedInterposerHops
\end{split}
\end{equation}

In a similar way, we modify the $ExpectedCongestionCost(r,j)$ in (\ref{eq:expected_routing_cost}) to account for the new interposer routing nodes.

The interposer-aware router lookahead is able to make routing decisions more quickly given the large delay difference of a path that crosses the interposer as opposed to a path that doesn't.

%   - Effectiveness / results
\subsection{Effectiveness of the enhancements}
\label{sec:CADeffect}
For all experiments in this paper, the remaining architecture parameters of the FPGA are taken from the ``flagship" architecture of the VTR project (\textit{k6\_frac\_N10\_mem32K\_40nm.xml}) \cite{luu2014vtr}. The parameters of this architecture are in line with both current commercial FPGAs and academic research into best practices. It consists of logic clusters with 10 fracturable 6-LUTs per block (N=10, k=6), and also includes 32kb RAM blocks and DSP blocks configurable to perform 9x9, 18x18 or 36x36 multipliers. The delay values in the architecture are taken from 40 nm circuit simulations and 40 nm commercial FPGAs. It uses unidirectional routing, with all wire segments having length L = 4.

We used VPR 7.0, with the enhancements we detailed above, and the architecture file \textit{k6\_frac\_N10\_mem32K\_40nm} in the experiments below. All experiments targeted the smallest FPGA (with number of rows equal to number of columns) which could accommodate a benchmark circuit; this represents a very full FPGA with little white space left, and hence presents a difficult case to an interposer-based FPGA as no die can be left mostly empty.

We have used the eight largest circuits from the VTR benchmark~\cite{vtr2012}, namely: \textit{bgm}, \textit{LU8PEEng}, \textit{LU32PEEng}, \textit{mcml}, \textit{mkDelayWorker32B}, \textit{stereovision0}, \textit{stereovision1} and \textit{stereovision2}. The size of these circuits ranges from $9,100$ to $153,000$ primitives (LUTs, FFs, etc.), with an average size of $52,600$ primitives. All results are the geometric mean over all circuits for a given interposer architecture.

To obtain the \textit{critical path delay} the circuits were run with a low stress routing with a channel width, $W = 1.3 \times minW$, where $minW$ is the minimum channel width for which the circuit is still routable.

\subsubsection{Effectiveness of placer optimizations}
\label{sec:effectiveness_of_placer_enhancements}
Enabling the enhancements described in Section~\ref{cad_enh_placer_subsection} improves the results .... haven't finished running these yet. So, <TODO EHSAN>: fill this section with actual results. For now, let's use a graph from Andre as a placeholder!
The graphs show that combining the timing with the bounding box optimizations led to an average of 20\% improvement on the area-delay product and the most significant contribution was made by incorporation of \textit{cut\_cost} into the wire cost function.

@@

\subsubsection{Effectiveness of router optimizations}
\label{sec:effectiveness_of_router_enhancements}
The router optimizations described in Section~\ref{sec:router_optimizations} greatly reduce the runtime of the router. Figure~\ref{fig:lookahead_runtime} shows the geomean of the router runtime for the 8 largest VTR circuits for different \textit{\% wires cut} values. On average, the router runs 31 times faster and finds identical minimum channel widths for all circuits.

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{router_opt.eps}
\caption{Router runtime with original and updated lookahead function.}
\label{fig:lookahead_runtime}
\end{figure}

The new router lookahead changes the priority of the routes discovered by the router. Using the new router lookahead function increases the critical path delay by 1.3\% on average as shown in Figure~\ref{fig:lookahead_fmax}. We consider this small quality loss acceptable and use the enhanced router lookahead in all our experiments due to its greatly reduced runtime.

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{router_opt_fmax.eps}
\caption{Impact of new vs. original router lookahead on the critical path delay.}
\label{fig:lookahead_fmax}
\end{figure}

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Adding Partitioning to the CAD Flow}
\label{sec:CADflow}
TODO: Javeed

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture results}
\label{resultsSection}

Using the best CAD settings found in Section~\ref{cadSection} and the best CAD flow described in Section~\ref{sec:CADflow}, we analyze the impact of the key architecture parameters: \textit{\% wires cut}, \textit{delay increase}, \textit{number of cuts}, and \textit{interposer-routing interface}. We run all the experiments using the setup detailed in Section~\ref{sec:CADeffect}.

%   - routing-interposer interface
\subsection{Routing-Interposer Interface}
As described in \ref{sec:interposer-routing-interface}, we have added three boolean options to VPR that allow us to experiment with different routing-interposer interfaces. Some interface choices provide more flexibility to route signals between different dice and hence help us achieve lower \textit{minimum channel width} and \textit{critical path delay}.

To measure the impact of routing-interposer interface parameters, we set \textit{delay increase}=$1ns$, \textit{number of cuts=1}, and \textit{\% wires cut=80\%}; then, we measure the impact of the \textit{bidirectional interposer wires}, \textit{fanin transfer}, and \textit{fanout transfer} parameters in different combinations. Figure~\ref{fig:all_arch_experiments} summarizes the results of these experiments; all results are normalized to the least flexible routing-interposer interface: one in which 80\% of wires end at the interposer and the other 20\% cross the interposer with the same directionality and routing switches they would have in a conventional FPGA.

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{arch_experiments3.eps}
\caption{Impact of routing-interposer interface flexibility on minimum channel width and circuit speed.}
\label{fig:all_arch_experiments}
\end{figure}

Figure~\ref{fig:all_arch_experiments} shows that using bidirectional interposer wires instead of unidirectional wires (column \textit{b} in Figure~\ref{fig:all_arch_experiments}) improves minimum channel width by 1\% and critical path delay by 0.5\%. Therefore, while it is beneficial to use bidirectional interposer wires, in isolation they do not dramatically improve performance.
